
---
title: "Problem Set 3 - Instrumental Variables"
author: 
- name: Author One
- name: Author Two
- name: Author Three
date: "24 March 2021"
output:
  html_document:
    highlight: tango
    theme: cosmo
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
    df_print: paged
---

```{r, include=FALSE}
# Keep this setup code chunk also in your problem set.
# Also, use this line to load the packages that are needed
library(tidyverse)
library(kableExtra)
library(wooldridge)
library(AER)
library(stargazer)

knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Learning goals

- Understanding the set up of IV studies
- Apply IV regression in practice
- Practicing data analysis


## Exercise 1

Use Google Scholar to find three examples of instrumental variables in published scientific research (in any area). List the three instruments describing (1-2 lines) the outcome variable, the endogenous treatment, and how the instrument solves the issue. 

Paper 1: The medial care costes of obesity: An instrumental variables approach
This paper explains the impact of obesity on medical costs by using *genetic variation in weight* as the instrumental variable (natural exp.). Weight is the endogenous treatment, which leads to a general underestimation of medical costs.The IV reduces error in weight reporting and distinguishes between natural obesity and obesity due to depression or injury. 

Paper 2: Economic Shocks and Civil conflict: An instrumental variables approach
The paper tests the impact of economic conditions (endogenous treatment) on the likelihood of civic conflict by using rainfall as the instrumental variable. They argue that *rainfall* correlates with economic growth (more or less rain (random) leads to more or less growth) which in turn is negatively related to civic conflict. Thus, a causal relationship is established.

Paper 3: Education and entrepreneurial choice: An instrumental variables approach
This paper wants to tackle the endogenous problem of the variable education and its impact on the decision to start a business. Since education is impacted by so many other variables (wealth, parents help, social class, place of living etc), they propose the instrumental variable *social class of parents*. Now the impact of education is much stronger, since students from lower classes are more likely to open a business. 

## Exercise 2 (no coding)

Install the package `wooldridge` and lead the data `bwght` (you can get help with `?bwght` after loading the package). Use the data to study the causal effect of smoking mothers on the weight of the babies. 

First think:   
1. What would be the ideal experiment to identify the relationship between maternal smoking and the weight of the baby? What would be the problem with that?

It would be ideal to randomly select pregnant women to smoke or not smoke during pregnancy and then weight all the babies after birth. Of course, we cannot do that because you cannot force pregnant women into smoking since smoking is really bad for the health of both the mother and the baby.

2. Can you think of a good instrument for smoking cigarettes? Better if the instrument... is not in the data! :)

IV: rainy and cold weather, since people tend to smoke more with sunny/warm weather. 
3. Describe the three requirement with respect to your instrument. 
1. causal effect: People tend to smoke more with sunny/ warm weather than by cold/ rainy.
2. Weather is always random (except if one believes the weather manipulation theories...which we don't)
3. Weather does not impact the weight of a baby, only the amount of cigarettes a pregnant women might smoke.

## Exercise 3
Find the variables for the number of cigarettes smoked during pregnancy and the weight of the babies (which is an indicator of overall health of newborn babies). 

### 3.1
Start by exploring the naive correlation between the number of cigarettes smoked per day during pregnancy on the birth weight in ounces: use the two variables to estimate a naive OLS regression model that predicts the weight of babies with the number of cigarettes smoked by the mother during the pregnancy. What do you find? is this an estimate of the causal effect of smoking? Why? Finally, choose from the variable list two potential omitted variables and add them to a second model. Is the estimate changing? Is this a causal effect now?

Model 1 shows a strong negative impact of cigarettes on baby's weight. The more cigarettes a mother smokes during pregnancy, the less weight a baby has after birth. No, this is no causal effect (only correlation), since smoking cigarettes is endogenous and so many other variables could have an impact on the baby's weight, for example *income of family* (if they are poor, maybe they do not have enough food and therefore the baby is light) or *gender of baby* might matter, since boys are generally a bit heavier than girls. 

Model 2 shows the results of the regression with omitted variables. However, the estimate is not changing. The two possible omitted variables are both significant but do not change the outcome of cigarettes and therefore are not omitted variables.

```{r}
data(bwght)
m1 <- lm(bwght ~ cigs, data= bwght)
summary(m1)

# model with 2 ommited variables
m2 <- lm(bwght ~  cigs + faminc + male, data= bwght)
summary(m2)

```

### 3.2
Check out the data and find a viable **instrument** for smoking. 
Defend the three requirements of the instrument.
IV: cigarette tax in home state
1. The more a mother has to be for a cigarette, the less likely she will buy them.
2. It is random, because the tax is different in each state and nobody would actually move into a different state just to have lower cigs. taxes.
3. The tax of cigarettes cannot influence a baby's weight directly.

### 3.3
Now manually compute the two stages of the 2SLS strategy (i.e., first stage, predict, and then use predicted values in the outcome equation).
```{r}
m3 <- lm (cigs ~cigtax, data = bwght)
pred <-predict(m3)


m4 <- lm (bwght ~ pred, data = bwght)
summary(m4)
```


### 3.4
Finally, estimate a 2SLS in one go calling the function `AER::ivreg()`. Self-learn how to use it. 
```{r}
m5 <- ivreg(bwght ~ cigs| cigtax, data = bwght)
summary(m5)
```


### 3.5 
Build a regression table with three models:  
1. `Naive OLS` with the simple naive regression of baby weights on cigarettes smoked;  
2. `IV manual` with the result from the two stages estimated manually (show only outcome equation);  
3. `IV-2SLS` with the result from `ivreg()`. 
```{r results='asis'}
stargazer(m1, m4, m5, type = "html")
```

Comment.

The strongest statistically significant relationship is definitely model 1 (without any instrument). When doing the 2sls approach manually, at least it stays slightly significant but with the ivreg() all significance vanishes. However, the constant is always significant, meaning that *no tax* increases the cigarette consumption and decreases the weight. Yet, the difference in taxes has no impact. 
Therefore, the analysis shows that with instrumental variable the impact of smoking mothers on a baby's weight vanishes dramatically, which makes sense, since there are many more reasons for a baby's weight - especially since the majority of mothers in the sample (1000+) do not smoke, but still have different baby weights.

## Exercise 4 

Skim through the paper by Acemoglu, Johnson, and Robinson (2001) [[link](https://www.aeaweb.org/articles?id=10.1257/aer.91.5.1369)]. Describe their research question, the source of endogeneity, and the instrumental variable that they use. (max 4 lines) Then, use the replication data stored [here](https://www.openicpsr.org/openicpsr/project/112564/version/V1/view?path=/openicpsr/112564/fcr:versions/V1/Acemoglu-Johnson-and-Robinson-datafiles&type=folder) to replicate Figure 1, Figure 3, and Table 4 (panel A only, p. 1386). 

**Warning**. You must work collaboratively using GitHub. Please create the repo at the class organization (not at your personal account). All the tips from PS1 still apply (plan the work, use issues, branches, pull requests...). 

Research Question: What are the fundamental causes of the large differences in income per capita across countries?
Source of endogeneity: very different colonization policies in different colonies, with different associated institutions
instrumental variable: mortality rates
They use mortality rates to explain differences in policies and institutions among european colonies, which then in turn influence up until today the income per capita across these countries. 

```{r}
fig1 <- haven::read_dta("Reply-to-Albouy-dataset-for-AER-replication-filing-May-2-2012.dta")


ggplot(fig1, aes( logmort0, loggdp)) +
  geom_text(aes(label=shortnam)) + geom_smooth(method = lm, se = FALSE) + labs(title = "FIGURE 1. REDUCED-FORM RELATIONSHIP BETWEEN INCOME AND SETTLER MORTALITY", x = "Log of Settler Mortality", y = "Log GDP per Capita") 


```
# figure 3
```{r}
ggplot(fig1, aes( logmort0, risk)) +
  geom_text(aes(label=shortnam)) + geom_smooth(method = lm, se = FALSE) + labs(title = "FIGURE 3. FIRST-STAGE RELATIONSHIP BETWEEN SETTLER MORTALITY AND EXPROPRIATION RISK", x = "Log of Settler Mortality", y = "Average Expropriation Risk") 


```
```{r results='asis'}
#base models
m1 <-ivreg(loggdp ~ risk | logmort0, data = fig1)
summary(m1)
m2 <- ivreg(loggdp ~ risk + latitude | logmort0 + latitude, data = fig1)
summary(m2)

#no neo-europ
ne <-filter( fig1, neoeuro == 0)
m3 <- ivreg(loggdp ~risk | logmort0, data = ne)
summary(m3)
m4 <- ivreg(loggdp ~risk + latitude | logmort0 + latitude, data = ne)
summary(m4)

# no africa
na <- filter(fig1, africa == 0)
m5 <- ivreg(loggdp ~risk | logmort0, data = na)
summary(m5)
m6 <- ivreg(loggdp ~risk  + latitude| logmort0 + latitude, data = na)
summary(m6)

#with continent
m7 <- ivreg(loggdp ~risk + africa + asia + other | logmort0 + africa + asia + other, data = fig1)
summary(m7)
m8 <- ivreg(loggdp ~risk + africa + asia + other + latitude | logmort0 + africa + asia + other + latitude, data = fig1)
summary(m8)

stargazer(m1, m2, m3, m4, m5, m6, m7, m8, type = "html")
```

